\part{Methods}
\label{pa:methods}
\chapter{Algorithms}
\section{Clustering}
To cluster or not to cluster is the question here. PPI networks already have
a great deal of edges and can be seen as clusters that we should not alter. On
the other hand, using clustering algorithms to make clusters out of
PPI-networks gives us the more control over the clusters. How big they should
be, how many of them we want, should we cluster on a certain attribute, or even
several? If we cluster the network, we have chosen to go with the \textit{MCL}
clustering algorithm in \textit{clusterMaker2}. We prototyped Ranklust with
\textit{Affinity Propagation}(AP), but a paper that performed a comparison
between MCL and AP concluded with MCL having the best performance in many
aspects when it came to PPI networks \cite{ap-vs-mcl}.

A drawback with clustering in CytoScape is that it directly alters the network.
However, MCL in clusterMaker2 does not create new edges between nodes. It is
keeping the ones that it qualifies as worthy to be in the cluster it is
currently creating. So it may disconnect, or delete, edges between nodes to
separate them in the clustering. This may not be feasible when ranking the
clusters afterwards, so the user should be able to choose between using
clusterMaker2's clustering algorithms, or just have clusters that already exists
through simple PPI connections between nodes.

\section{Ranking clusters}
We have used five algorithms to rank the clusters, MAA, MAM, PR, PRWR and HITS.
The two first are our own simple methods, used for prototyping, but PR is a
proven ranking algorithm for PPI-networks in cancer. It is our belief that the
PRWR results have the highest likeliness of giving successfull results, should
ranking of clusters unveil new information on network biomarkers for prostate
cancer.

Every algorithm implemented in Ranklust for ranking clusters except one takes
node and/or edge scores as input for calculating the scores for each node.

\subsection{Single Attribute Additive Method (SAA)}
Single attribute additive method is the simplest of the four algorithms. It
gives the user the option to choose a single node attribute to score the
clusters on. The attribute is required to be a number of some sort, but that is
the only constraint.

It goes through all of the nodes in each cluster and sums up the number-
attribute the user chose. Each cluster is then ranked based on the accumulated
sum and ranked descending, with the highest ranking cluster as the most likely
prostate cancer biomarker.

\subsection{Multiple Attribute Additive Method (MAA)}
Multiple attribute additive method is the next step. This opens up the
opportunity for the user to use an unlimited number of number-attributes on both
nodes and edges. The part after that is exactly the same as in the SAA
algorithm. It goes through all of the nodes in each cluster and sums up the
attributes for each cluster.

There is a question as to how to rank the edges in the cluster. We chose to rank
each edge as it is listed to the user in CytoScape. So if it is listed only once
time in the edge table, it will only be scored additively once. This decision
was based on simplicity. To not represent the edge as something the user did not
define it as, or is unable to understand.

\subsection{Multiple Attribute Multiplication Method (MAM)}
Multiple attribute multiplication method is to some degree redundant,
considering what exists from before in Ranklust. The only difference from MAA is
the scale the scores will be in. MAM adds the scores from each node and edge in
the cluster through multiplication, MAA does it with addition. 

Another detail to the MAM is that it will ensure that the scores either increase
or stay the same, it will never decrease. So when values that are below one is
combined in MAM, it will gain a plus one score. In other words, 0.23 times % format!
0.5 will result in 1.23 times 1.5. The same is true for situations where only
one of the numbers is below one. 0.23 times 1.5 becomes 1.23 times 1.5. Zero
scores will get ignored.

\subsection{Personalized PageRank}
A Random Walks algorithm which used priors, called seed-weighted random walks
ranking \cite{sw-rwr}, proved to be effective at prioritizing biomarker
candidates. PageRank is a Random Walks based algorithm. PageRank was previously
used by Google to rank webpages \cite{pagerank}. Personalized PageRank is a
modified version of PageRank, where nodes and edges can be assigned a score
prior to the PageRank's traversal of the network. 

The difference between MAA and MAM compared to PRWP is how the network is
scored. MAA and MAM calculates the score for each cluster by summing up the
attributesin edges and nodes. PRWP scores the current network, regardless of the
clustering attribute. This enables the user to analyze the network as-is before
clustering and let the clustering just stand for how the scores are grouped. The
second option the user has is to let the clustering algorithm create a new
network based on the clusters, and then rank the network. The last option gives
the clustering algorithm a bigger impact on the score. Though, this is only true
for PageRank related algorithms in Ranklust like PR, PRWP and HITS. Because they
rely on an alpha variable controlling the probability of a reset in the
traversal. This "traversal reset" is automatically instantiated if the algorithm
hits a node with no outgoing edges. As the clustering algorithms might change
the edges in the network, the outcome of performing ranking on a cluster-created
network, or a network with only a cluster attribute to identify the clusters,
can be potentially vastly different. % Insert pictures!

\subsection{Hyperlink-Induced Topic Search}
Hyperlink-Induced Topic Search, HITS, is an algorithm that is similar to
PageRank and was developed around the same time \cite{hits}\cite{hits-origin}.
HITS will not be used to calculate cluster scores and ranks, due to the fact
that it does not require any form of weighting in nodes or edges. Though, it can
be used in combination with other ranking algorithms through the use of MAM or
MAA. 

An example of this could be running PRWP with a score attribute on a
network, then HITS on the same network. The next step would be to combine the
two scores from PRWP and HITS with MAA. Though, the idea of achieving novel
information on network biomarkers for prostate cancer through this workflow is
pure speculation.

\chapter{Comparison to other biomarker identification methods}
To compare the ranking of biomarker clusters versus ranking of single biomarker
nodes, % continue here when you know the answer

\chapter{Creation of Neo4J database} % Insert appendix links to the scripts!
\section{Data preparation}
We created a Neo4J database and used a slightly modified CytoScape app to
communicate with it. Input to the database is a modified version of data. The
data started as a PPI file from the StringDB. The file had a size of 3,9GB and
consisted of interactions between two proteins. Each line had two Ensembl
Protein IDs (ENSP), the type of interaction, what direction the interaction went
in and a score for the interaction. 

\begin{verbatim}
item_id_a	item_id_b	mode	action	a_is_acting	score
9606.ENSP00000000233	9606.ENSP00000000233	binding		0	800
\end{verbatim}

Then we strip this file in order to get a file with just the ENSP IDs. One line
per ID.

\begin{verbatim}
ENSP00000000233
ENSP00000005340
\end{verbatim}

Next up is the conversion from protein IDs to gene IDs coupled with Entrez ID.
This information we got from UniProt through the stripped ENSP IDs. The file
with the data is formatted like this:

\begin{verbatim}
Entry	EnsembleID	    Gene names
P84085	ENSP00000000233	ARF5
\end{verbatim}

Then match the \textit{EnsembleID} from UniProt, which is a ENSP ID, to the
\textit{item\_id\_a} and \textit{item\_id\_b} data from StringDB (without the
9606.  prefix). Then combine the Entrez id, which is the \textit{Entry} from
UniProt with \textit{Gene names} with a square (\textit{\#}) between them as
a separator to make it easier to split them when creating Cypher and GraphML
import queries for Neo4J. This combination makes up a single gene, and each line
in the file we write this information to consist of two genes and a type of
binding between them. The type of binding is provided by the StringDB file
mentioned first in this section. % label and refer instead of mentioning it!

\section{Neo4J setup and data import}

\subsection{Setup}
Not much of a setup is needed to get the database up and running. Every setting
that was changed in testing and using the database was set in the
\textit{conf/neo4j-server.properties} file, located in the directory relative to
the Neo4J database installation.

The database
location does not need to be set up, but knowing the location of the data that
is saved in the Neo4J database is somewhat crucial when working with several
database instances.

\begin{verbatim}
org.neo4j.server.database.location=data/string_mini.db
org.neo4j.server.webserver.port=7474
\end{verbatim}

NB! That path is relative to the location the neo4j database is run from.

Turning authentication on/off is also smart for testing purposes. Communicating
with a Neo4J database instance that demands authentication also requires the
usage of the modified CytoScape app \textit{CyNeo4J}. This is because the app
does support everything we need except from authenticating with a database. For
the database, this is just having a boolean value \textit{true}/\textit{false}.

\begin{verbatim}
dbms.security.auth_enabled=true
\end{verbatim}

\subsection{Import}
Importing data into Neo4J is not very mature at this point. You have the
possibility to just use Neo4J's query language, \textit{Cypher}. However,
importing the whole gene data from StringDB and UniProt combined takes a long
time in Cypher, even if the language is the native language to Neo4J. The
alternative we used was \textit{neo4j-shell-tools} \cite{neo4j-tools}. It is
easy to install and supports import with \textit{CSV}, \textit{Geoff} and
\textit{GraphML} filetypes. A GraphML import took under 1 minute on my old and
underperforming laptop with 4GB memory, 2GHz CPU and a 128GB SSD \cite{laptop}.
In comparison, the Cypher import took several hours, with the same amount of
data. Testing related to Neo4J was not done with CSV-files, because CytoScape
does not support exporting data of a whole network to a CSV file, only partial
information. On the other hand, CytoScape has great support for exporting data
of a whole network to a GraphML file.

The Python scripts used to initialize the Neo4J database with data create either
GraphML or Cypher. Importing any of them result in the same type of data in the
database, but because of the heavily increased speed using GraphML over Cypher,
we used GraphML exclusively to initialize the database. The nodes are created
with the forementioned \textit{Entrez ID} as primary key and \textit{Gene name}
as the displayed name in the Neo4J GUI.

\chapter{CytoScape implementation}
The clustered network can be constructed with an AP, short for Affinity
Propagation, clustering algorithm \cite{affinity-propagation}. Affinity
propagation clustering concentrates on the nodes in the network that are binding
the rest of them together. Using affinity propagation for clustering will
produce results that represent a grouping of nodes that are coupled by seemingly
unimportant nodes to most clustering algorithms. But the AP algorithm is good at
expressing nodes that are not highly connected to many nodes, but rather the
nodes that are binding other highly connected nodes together. This results in
bigger clusters that can be a target for methods that cure cancer by severing
the interaction between biomarker genes.

It is also discussed how AP performs versus Markov clustering \cite{ap-vs-mcl}.
And since Markov algorithm performs better on protein interaction, it will also
be used to cluster the networks. An analysis between the rankings that come from
the Markov and AP clusterings will be performed, which hopefully will give
concrete results as to the pro's and con's of each algorithm. Some questions
should be raised as the analysis is done. For example, is both of the algorithms
good, but have different uses, even if they are not directly involved with the
ranking done after the clustering?  Are either AP or Markov useless for
a particular type of ranking afterwards?

The information provided through the whole process from what type of nodes
(protein or gene), interaction between the nodes and what kind of biomarker we
want to identify are all factors that will have a great impact on how all of
this should be combined. The order of operations on the network will also effect
the result. For example, AP clustering may create a few big clusters and many
small ones. At this stage, the results will consist of the biggest clusters
constructed from AP clustering that has focus on pure connection between nodes
and not their attributes. At this point, the cluster ranking algorithm of
Ranklust will be run in order to produce a picture of potential biomarkers. This
picture is the first and simplest step that will be used as a result for an
analysis. The analysis in this thesis will focus on validating the cluster
ranking scores as ways of indicating potential biomarkers. 
