\documentclass[a4paper,UKenglish]{ifimaster}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{duomasterforside}
\usepackage[T1]{fontenc, url}
\usepackage{graphicx,babel,csquotes,textcomp,varioref,float}
\usepackage[backend=biber,style=numeric-comp,sorting=nyt,maxbibnames=20,maxcitenames=20]{biblatex}
\usepackage[toc]{glossaries}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{parskip}
\usepackage{fancyvrb}
\usepackage{tabularx}
\usepackage{rotating}
\usepackage{multirow}
\urlstyle{sf}
\graphicspath{{images/}}
\addbibresource{macros.bib}
\addbibresource{references.bib}

\lstset{
    breaklines=true,
    frame=tb,
    numbers=left,
    stepnumber=1,
    numberstyle=\color{gray},
    keywordstyle=\color{blue},
    stringstyle=\color{red},
    commentstyle=\color{olive},
    keepspaces=true
}

\makeglossaries
\newglossaryentry{golden}
{
    name=golden standard,
    description={The prior scoring standard evolved from the DisGeNET and Dragon
    Database of Genes associated with Prostate Cancer}
}
\newglossaryentry{maa}
{
    name=MAA,
    description={The Multiple Attribute Additive method described to score
    clusters}
}
\newglossaryentry{mam}
{
    name=MAM,
    description={The Multiple Attribute Multiplicative method described to score
    clusters}
}
\newglossaryentry{pr}
{
    name=PR,
    description={The PageRank method described to score clusters}
}
\newglossaryentry{prwp}
{
    name=PRWP,
    description={The PageRank With Priors method described to score clusters}
}
\newglossaryentry{hits}
{
    name=HITS,
    description={The Hyperlink-Induced Topic Search method described to score
    clusters}
}
\newglossaryentry{maven}
{
    name=Maven,
    description={The Java build tool used to compile and build clusterMaker2 and
    Ranklust}
}
\newglossaryentry{ide}
{
    name=Integrated Development Environment,
    description={A development environment specialized in developing software
    products}
}
\newglossaryentry{mcl}
{
    name=MCL,
    description={Markov Cluster algorithm used in Cytoscape to cluster the
    networks}
}
\newglossaryentry{rsquared}
{
    name=R-squared,
    description={The coefficient of determination when talking about the grade
    of fitness of a linear regression}
}
\newglossaryentry{jensen}
{
    name=DISEASE,
    description={DISEASE database used for retrieving z-values, p-values and
    manually curated disease-gene associations}
}
\newglossaryentry{pipeline}
{
    name=pipeline,
    description={The time used from start to end when analyzing the network in
    Cytoscape}
}
\newglossaryentry{dragon}
{
    name=DDPC,
    description={Dragon Database of Genes associated with Prostate Cancer}
}
\newglossaryentry{movember}
{
    name=potential prostate cancer genes,
    description={Data retrieved from the Movember Prostate Cancer Project}
}

\title{Ranklust}
\subtitle{An extension of the Cytoscape clusterMaker2 plugin and its application
to prioritize network biomarkers in prostate cancer}
\author{Henning Lund-Hanssen}

\begin{document}
\duoforside[program={Programming and Networks},
    dept={Department of Informatics},long]

\frontmatter{}

\setlength{\parskip}{12pt}
\setlength{\parindent}{12pt}

\chapter*{Abstract}
Prostate-specific antigen (PSA) is a prostate cancer biomarker used for
screening that in a great number of cases has led to overtreatment. The main
flaw with this biomarker is that mutation or abnormality of a single gene is
seldom the single cause of a disease. Rather, it is the process of interactions
between several components in a complex network. Next-generation technologies
provide bioinformaticians with data on a scale worthy of being labeled "Big
Data". One of the advantages gained is the potential to identify and prioritize
genes that promote diseases by analyzing biologically related networks. Another
advantage is the newly gained understanding of the topological organization of
large-scale molecular networks, which has been achieved through clustering
networks, creating subnetworks representing functions. The goal of this
assignment was to combine these two newly adapted methods in bioinformatics to
develop a plugin for Cytoscape, a network data integration, analysis and
visualization tool, in order to prioritize network biomarkers for prostate
cancer. Through cross-validation, benchmarking against text mined, manually
knowledge curated and experimental data, Ranklust has demonstrated this ability
by using the PageRank with Priors algorithm to rank clusters made with the
Markov Cluster algorithm in a protein-protein interaction network (PPI). It
ended up being merged to one of the future branches of clusterMaker2 on GitHub.

\chapter*{Acknowledgements}
I would first like to thank my two thesis supervisors. Research Fellow and
supervisor Trevor Clancy of the Institute for Cancer Research at Oslo University
Hospital and Group leader and co-supervisor Eivind Hovig of the Institute of
Cancer Research at the Oslo University Hospital and Professor II of the
Institute of Informatics at the University of Oslo. You guys have always been
available when I needed advice, supported me throughout the whole process and
provided me with invaluable knowledge.

Finally, I must express my very profound gratitude to my family, for their
unfailing support and continuous encouragement. To my fellow study companions at
room 3417 Assembler "Assa", Ole-Johan Dahls hus, over the years you guys have
taught me that procrastination is an art best performed in great company. Thank
you.

\tableofcontents{}
\listoffigures{}
\listoftables{}
\lstlistoflistings
\printglossaries

\mainmatter{}

\include{intro}
\include{methods}
\include{results}
\include{conclusion}
\backmatter{}
\printbibliography
\end{document}
